---
title: "Practical Machine Learning - Course Project"
author: "Giovanna14"
date: "22/03/2015"
output: html_document
---

##Synopsis##

This project is based on the Weight Lifting Exercise Dataset 
available at http://groupware.les.inf.puc-rio.br/har. 
The dataset is part of a human activity recognition research
making use of wearable accelerometers to determine how well
a given activity is performed and automatically detect execution
mistakes. The goal of this project is to to use this dataset to build 
a machine learning algorithm that predicts the manner in which the 
participants executed the exercise.

##Dataset##

Data were collected from
accelerometers on the belt, forearm, arm, and dumbbell of 6 people,
while they were lifting a dumbbell in 5 different ways, a correct 
one and 4 different wrong ways, simulating typical mistakes.
The variable `classe` in the dataset defines the way how the exercise
was performed. It takes up 5 possible values, A for
correct excution and B, C, D, E for each of the 4 simulated 
execution mistakes. 

```{r dataset,cache=TRUE}
#downloading data
URL1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
URL2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
file1 <- "pml-training.csv"
file2 <- "pml-testing.csv"
if (!file.exists(file1)){download.file(URL1,file1,method="curl")}
if (!file.exists(file2)){download.file(URL2,file2,method="curl")}
```
A preliminary inspection of the pml-training.csv file shows 
the presence of several variables whose values are almost only NAs, 
blank values, or "#DIV/0!". So while loading the data into data frames
we redifine all these values as NAs.

```{r loading, cache=TRUE}
training <- read.csv(file1,na.strings=c("NA","","#DIV/0!"))
testing <- read.csv(file2,na.strings=c("NA","","#DIV/0!"))
```

##Preparatory steps##

For our purposes we use the `caret` package, so we load this and
other additional packages that are required subsequently for modelling 
and plotting purposes.

```{r packages}
library(ggplot2);library(lattice);library(caret)
library(parallel); library(doParallel) #for parallel processing
library(AppliedPredictiveModeling)
library(randomForest);library(rpart);library(partykit)
```

Since there is a large number of observations, we can split the provided
training set into three subsets, a training, 
a validation, and a testing set approximately 
comprising 60%, 20%, and 20% of the original set, respectively. 
Since there are several columns that contains almost only NAs,
implementing an imputing strategy for these columns does not seem
appropriate, hence we discard them from our newly created 
training set. In so doing, we are automatically discarding the
columns containing computed statistical quantities (e.g. mean, kurtosis,
skewness), with the exception of total accelerations.  
We also discard the first 7 columns that do not
contain useful information for the purpose of modelling (sequential 
numbers, names of participants, time stamps, windows for measurement
collection).

```{r partitioning and cleaning,cache=TRUE}
## split original training set into a training, a validation and
## a test set
set.seed(1521)
inTrain <- createDataPartition(y=training$classe,p=0.6,list=FALSE)
trainset <- training[inTrain,]
notrain <- training[-inTrain,]
inVal <- createDataPartition(y=notrain$classe,p=0.5,list=FALSE)
validset <- notrain[inVal,]
testset <- notrain[-inVal,]
# remove first 7 columns that are not useful for modelling
trainset <- trainset[,-c(1:7)]
# remove columns with NAs
trainset <- trainset[,colSums(is.na(trainset))==0]
dim(trainset)
```

##Exploratory analysis##

We carry out some exploratory analysis on the training set to
identify possible trends and outliers.

```{r pairs,cache=TRUE}
featurePlot(x=trainset[,c("roll_belt","pitch_forearm",
                          "roll_dumbbell","roll_forearm",
                          "accel_dumbbell_z")],y=trainset$classe,
            plot="pairs")
```

There are some outliers, and one would have to investigate them.
However, in the following we proceed without any data processing.

##Building predictive models##

In the following we use `train` from the `caret` package to build
predictive models with different subsets of predictors and with
two alternative models, i.e. the tree-based model `rpart` and 
the random forest model `rf`.

###Tree-based models###

We create an rpart model with all predictors in the cleaned training set.

```{r rpart with all predictors,cache=TRUE}
set.seed(3528)
registerDoParallel(clust <- makeForkCluster(3))
fit1_rpart <- train(classe ~ ., data=trainset,method="rpart",
            trControl=trainControl(method="cv",number=5,
                                     allowParallel=TRUE) 
            )
stopCluster(clust)
fit1_rpart$finalModel
# importance of variables
varImp(fit1_rpart)
```

The resulting classification tree is shown in Figure 1.

```{r rpart model trees,cache=TRUE}
fit1party <- as.party(fit1_rpart$finalModel)
plot(fit1party,main="rpart model - all predictors")
```

**Figure1.** Model trees obtained using an rpart model with all 
predictors.

### Random forest models###

For random forest models we first use all predictors in the 
cleaned training set and then we try to progressively reduce the
number of predictors.

```{r random forest with all predictors,cache=TRUE}
# use parallel processing on 3 CPUs to train with random forest with all predictors
set.seed(3528)
registerDoParallel(clust <- makeForkCluster(3))
fit1_rf <- train(classe ~ ., data=trainset,method="rf",ntree=200,
            trControl=trainControl(method="cv",number=5,
                                     allowParallel=TRUE) 
            )
stopCluster(clust)
fit1_rf$finalModel
# importance of predictors
imp1_rf <- varImp(fit1_rf)
varImp(fit1_rf)
# create a subset of the training set retaining only the most
# important predictors
mostImpVars1_rf <- rownames(imp1_rf$importance)[imp1_rf$importance>10]
subtrainset1_rf <- trainset[,c(mostImpVars1_rf,"classe")]
dim(subtrainset1_rf)
```

We now use only the most important predictors identified in the previous step.

```{r random forest with most important predictors,cache=TRUE}
# use parallel processing to train with random forest
set.seed(3528)
registerDoParallel(clust <- makeForkCluster(3))
fit2_rf <- train(classe ~ ., data=subtrainset1_rf,method="rf",ntree=200,
            trControl=trainControl(method="cv",number=5,
                                     allowParallel=TRUE) 
            )
stopCluster(clust)
fit2_rf$finalModel
imp2_rf <- varImp(fit2_rf)
varImp(fit2_rf)
# create a subset of the training set retaining only the most
# important predictors
mostImpVars2_rf <- rownames(imp2_rf$importance)[imp2_rf$importance>5]
subtrainset2_rf <- trainset[,c(mostImpVars2_rf,"classe")]
dim(subtrainset2_rf)
```

We further reduce the number of predictors, based on the most
important predictors identified in the previous step.

```{r random forest further reduced predictors,cache=TRUE}
# use parallel processing to train with random forest
set.seed(3528)
registerDoParallel(clust <- makeForkCluster(3))
fit3_rf <- train(classe ~ ., data=subtrainset2_rf,method="rf",ntree=200,
            trControl=trainControl(method="cv",number=5,
                                     allowParallel=TRUE) 
            )
stopCluster(clust)
fit3_rf$finalModel
imp3_rf <- varImp(fit3_rf)
varImp(fit3_rf)
```

###Comparing and fine-tuning random forest models###

Then we have a look at the random forest model error rates.

```{r model errors,cache=TRUE}
par(mfrow=c(1,3))
plot(fit1_rf$finalModel,type="l",
     main="Error rates - Model 'fit1_rf'")
legend("topright",colnames(fit1_rf$finalModel$err.rate),
      col=1:6,cex=0.8,fill=1:6)
plot(fit2_rf$finalModel,type="l",
     main="Error rates - Model 'fit2_rf'")
legend("topright",colnames(fit2_rf$finalModel$err.rate),
      col=1:6,cex=0.8,fill=1:6)
plot(fit3_rf$finalModel,type="l",
     main="Error rates - Model 'fit3_rf'")
legend("topright",colnames(fit3_rf$finalModel$err.rate),
      col=1:6,cex=0.8,fill=1:6)
par(mfcol=c(1,1))
```
**Figure 2.** Error rates for our three different rf models as a function
of the number of trees.

We apply the different models to the validation set and look at 
the achieved accuracy, to choose our final model.

```{r predicting,cache=TRUE}
# look at confusion matrices on validation set
predfit1_rp_valid <- predict(fit1_rpart,validset)
predfit1_rf_valid <- predict(fit1_rf,validset)
predfit2_rf_valid <- predict(fit2_rf,validset)
predfit3_rf_valid <- predict(fit3_rf,validset)
fit1_rp_CM <- confusionMatrix(predfit1_rp_valid,testset$classe)
fit1_rf_CM <- confusionMatrix(predfit1_rf_valid,testset$classe)
fit2_rf_CM <- confusionMatrix(predfit2_rf_valid,testset$classe)
fit3_rf_CM <- confusionMatrix(predfit3_rf_valid,testset$classe)
print(c("rpart model",fit1_rp_CM))
print(c("random forest model1",fit1_rf_CM))
print(c("random forest model 2",fit2_rf_CM))
print(c("random forest model 3",fit3_rf_CM))
```

The rpart model has very low accuracy and we do not consider it
any further. The accuracy of the random forest models decreases
when reducing the number of predictors, but with 
`r dim(subtrainset2_rf)[2]-1` predictors still has $\sim$ 98% accuracy,
as evaluated on the validation set.
although with an increased error rate for `classe` = B.
We feel that we can retain this last model, that is slightly less
accurate but simpler than the one which uses all predictors and 
reduces the risk of overfitting.

##Final model##

Based on the results from the use of the validation set
we choose as our final model `fit3_rf`. This was obtained with the
random forest method with 200 trees and 5-folds cross-validation
by using the `r dim(subtrainset2_rf)[2]-1` predictors that were 
identified as the most important ones. 

We apply this final model to our test set to evaluate the
out-of-sample error.

```{r prediction and confusion matrix on testset,cache=TRUE}
predfit3_rf_test <- predict(fit3_rf,testset)
confusionMatrix(predfit3_rf_test,testset$classe)
```

Finally, we use our chosen model to make predictions 
on the provided testing dataset and save them to a text
file for the submission part
of this assignment.

```{r sumission part,cache=TRUE}
pred_final <- predict(fit3_rf,testing)
answers <- as.character(pred_final)
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(answers)
```